### 2023-06-08 paper/localpsf.tex
grammar, typos, and other such edits

%Existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, which we cannot do. 
We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applies. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of the support $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. The moment method that we use estimates the support of $\phi_x$ more accurately than random noise probing in resolution analysis, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
In this work, we form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. 
%These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BorziSchulz11,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal to each other under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the ``regularization Hessian'') which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
hence more informative data leads to larger eigenvalues and therefore a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information---components of the parameter that could be learned from components of the observations that have a large signal to noise ratio are instead reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament---if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,HartlandEtAl23}. 
Although the required number of matrix-vector products grows nearly linearly in the discretized parameter dimension and linearly in the rank of the off-diagonal blocks, the required number of matrix-vector products is still large (e.g., hundreds to thousands).
%These methods are asymptotically scalable in the sense that the required number of matrix-vector products grows nearly linearly in the dimension of the discretized parameter and linearly in the rank of the off-diagonal blocks. But despite asymptotic scalability, from a practical perspective the required number of matrix-vector products is large (e.g., hundreds to thousands). 
%Improving the peeling process to reduce its cost is currently an area of active research~\cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently an area of active research (see, e.g.,~\cite{LevittMartinsson22}), so more efficient matrix-free H-matrix methods may be discovered in the future. 
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator actions, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. Not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(w) := \int_\Omega \int_\Omega w(y) \Aker(y,x) u(x) \mathrm{d}x \mathrm{d}y.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$.
The integral kernel, $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$, exists in principle but is not easily accessible. In this section we describe how to extend the action of $\Aop$ to distributions, which allows us to define impulse responses (Section~\ref{sec:impulse_response_2}), we then state the conditions on $\Aop$ that the proposed method requires (Section~\ref{sec:conditions_2}), and detail finite element discretization (Section~\ref{sec:finite_element_kernel}).

@@

The action of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Let $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well-defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Aker(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is acceptable if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$, as long as it is non-negative for the vast majority of pairs of points $(y,x)$, and as long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error, and depending on the severity of the violation, the proposed method may fail.
%Not all operators of interest (and in particular not all Hessians) satisfy these conditions, but many operators of practical interest do satisfy these conditions, and our method is highly effective for approximating these operators.

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / \spatialvol(x_i).
\end{equation*}
We compute the \emph{impulse response batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*
	=\sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / \spatialvol(x_i).
\end{equation}
The last equality in~\eqref{eq:dirac_comb_H_action} follows from linearity and the definition of $\impulseresponse_{x_i}$ in~\eqref{eq:impulse_response_delta_action}. Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) =
	\begin{cases}
		\combresponse_k(z) \spatialvol(x_i), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, $\xi_k \mapsto \left(\Aop \diraccomb_k^*\right)^*$, we recover $\impulseresponse_{x_i}$ for every point $x_i \in \pointbatch_k$.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit). The 
%number of impulse response batches used to build these 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. Then, H-matrix approximations of the Gauss-Newton Hessian, 
$$\bm{H}_\text{gn}\approx \widetilde{\bm{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\bm{H}}$ to be (approximately) symmetric positive semi-definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel. The numerical results show that~$\widetilde{\bm{H}}$ is a good preconditioner for both~$\bm{H}_\text{gn}$ and~$\bm{H}$. 
%The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%

>>>>

%Existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, which we cannot do. 
We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applies. The impulse response, $\phi_x$, associated with a point, $x$, is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of the support $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. The moment method that we use estimates the support of $\phi_x$ more accurately than random noise probing in resolution analysis, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
In this work, we form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. 
%These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BorziSchulz11,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal to each other under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the \emph{regularization Hessian}) which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
hence more informative data leads to larger eigenvalues and therefore a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information. If the regularization is too strong, then components of the parameter that could be learned from the observations
%components of the observations that have a large signal to noise ratio 
would instead be reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament: if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,HartlandEtAl23}. 
Although these methods are promising,
%required number of matrix-vector products grows logarithmically in the discretized parameter dimension and linearly in the rank of the off-diagonal blocks, 
the required number of operator applies is still large (e.g., hundreds to thousands).
%These methods are asymptotically scalable in the sense that the required number of matrix-vector products grows nearly linearly in the dimension of the discretized parameter and linearly in the rank of the off-diagonal blocks. But despite asymptotic scalability, from a practical perspective the required number of matrix-vector products is large (e.g., hundreds to thousands). 
%Improving the peeling process to reduce its cost is currently an area of active research~\cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently an area of active research (see, e.g.,~\cite{LevittMartinsson22}), so more efficient matrix-free H-matrix methods may be discovered in the future. 
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator applies, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. Not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(w) := \int_\Omega \int_\Omega w(y) \Aker(y,x) u(x) \mathrm{d}x \mathrm{d}y.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$.
The integral kernel, $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$, exists in principle but is not easily accessible. In this section we describe how to extend the application of $\Aop$ to distributions, which allows us to define impulse responses (Section~\ref{sec:impulse_response_2}), we then state the conditions on $\Aop$ that the proposed method requires (Section~\ref{sec:conditions_2}), and detail finite element discretization (Section~\ref{sec:finite_element_kernel}).

@@

The application of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Let $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the application of $\Aop$ on the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this operator application by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well-defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Aker(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is acceptable if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$ so long as it is non-negative for the vast majority of pairs of points $(y,x)$, and so long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error, and depending on the severity of the violation, the proposed method may fail.
%Not all operators of interest (and in particular not all Hessians) satisfy these conditions, but many operators of practical interest do satisfy these conditions, and our method is highly effective for approximating these operators.

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / \spatialvol(x_i).
\end{equation*}
We compute the \emph{impulse response batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*
	=\sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / \spatialvol(x_i).
\end{equation}
The last equality in~\eqref{eq:dirac_comb_H_action} follows from linearity and the definition of $\impulseresponse_{x_i}$ in~\eqref{eq:impulse_response_delta_action}. Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) =
	\begin{cases}
		\combresponse_k(z) \spatialvol(x_i), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By applying the operator once, $\xi_k \mapsto \left(\Aop \diraccomb_k^*\right)^*$, we recover $\impulseresponse_{x_i}$ for every point $x_i \in \pointbatch_k$.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit). The 
%number of impulse response batches used to build these 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. Then, H-matrix approximations of the Gauss-Newton Hessian, 
$$\bm{H}_\text{gn}\approx \widetilde{\bm{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\bm{H}}$ to be (approximately) symmetric positive semi-definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\bm{\preconditioner}$
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel. The numerical results show that~$\widetilde{\bm{H}}$ is a good preconditioner for both~$\bm{H}_\text{gn}$ and~$\bm{H}$. 
%The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%
